API design for machine learning software: experiences from the scikit-learn project

### 2013-06-26 paper.tex
Some formulation changes

\begin{description}
  \item[Consistency.]
       All objects (basic or composite) share a consistent interface composed of
       a limited set of methods. This interface is documented in a consistent
       manner for all objects.
  \item[Inspection.]
       Constructor parameters and parameter values determined by learning
       algorithms are stored and exposed as public attributes.
  \item[Non-proliferation of classes.]
       Learning algorithms are the only objects to be of custom classes.
       Datasets and hyper-parameter names and values are represented as standard
       Python, NumPy or SciPy classes rather than new ones whenever feasible.
       This keeps \sklearn easy to use and easy to combine with other libraries.
  \item[Composition.]
       Many machine learning tasks are expressible
       as sequences or combinations of transformations to data.
       Some learning algorithms are also naturally viewed
       as meta-algorithms parametrized on other algorithms.
       Whenever possible, such algorithms are implemented and composed from
       existing building blocks.
  \item[Sensible defaults.]
       Whenever an operation requires a user-defined parameter,
       a default and appropriate value is defined by the library.
       The default value should cause the operation to be performed
       in a sensible way (giving a baseline solution for the task at hand).
\end{description}

@@

In \sklearn, we chose a representation of data that is as close as
possible to the matrix representation: datasets are encoded as NumPy
multidimensional arrays for dense data and as SciPy sparse matrices for sparse
data. While these may seem rather unsophisticated data representations when
compared to more object-oriented constructs, such as the ones used by
Weka \citep{hall2009weka}, they bring the prime advantage of allowing us to rely
on efficient NumPy and SciPy vectorized operations while keeping
the code short and readable.  This design choice has also been motivated by
the fact that, given their pervasiveness in many other scientific Python
packages, many scientific users of Python are already familiar with NumPy dense
arrays and SciPy sparse matrices.
From a practical point of view, these formats also provide a bunch of
data loading and conversion tools which make them very easy to use in many
contexts. Moreover, for tasks where the inputs are text files or semi-structured
objects, we provide \textit{vectorizer} objects that efficiently convert such
data to the NumPy or SciPy formats.

@@

The \textit{estimator} interface is at the core of the
library. It defines instantiation mechanisms of objects and exposes a
\texttt{fit} method for learning a model from training data.  All supervised and
unsupervised learning algorithms (e.g., for classification, regression or
clustering) are offered as objects implementing that interface. Machine
learning tasks like feature extraction, feature selection or dimensionality
reduction are also provided as estimators.

@@

clf = LogisticRegression(penalty="l1")
clf.fit(X_train, y_train)
\end{pythoncode}
In this snippet, a \texttt{LogisticRegression} estimator is first initialized by
setting the \texttt{penalty} hyper-parameter to \texttt{"l1"}, hence enabling
$\ell_1$ regularization. Other hyper-parameters (such as \texttt{C}, which
defines the strength of the regularization) are not explicitly defined and are
thus set to the default values. Upon calling \texttt{fit}, the model is then
learned from the training arrays \texttt{X\_train} and \texttt{y\_train},
and stored within the object for later use.
%% (In this example, \texttt{fit} is called solely for its side effects.)
Since all estimators share the same interface, using a different learning algorithm is
as simple as replacing the constructor (the class name).
For example, to build a random forest on
the same training data, one would simply have to replace
\texttt{LogisticRegression(penalty="l1")} in the snippet above by \\
\texttt{RandomForestClassifier()}.

In \sklearn, classical learning algorithms are not the only objects to be
implemented as estimators. For example, preprocessing routines (e.g., scaling of
features) or feature extraction techniques (e.g., vectorization of text
documents) also implement the same interface. Even processing steps that are
entirely stateless and therefore do not require the \texttt{fit} method to
perform any useful work implement the estimator interface. As we will illustrate
in the next sections, this design pattern is indeed of prime importance for
consistency, composition and model selection reasons.

@@

Building on the core interface introduced in the previous section, we now
present advanced API mechanisms for building meta-estimators,
composing complex estimators and then selecting models. We also discuss design
choices allowing for an easy use and extension of \sklearn.

@@

A distinguishing feature of the \sklearn API is its ability to
compose new estimators from several base estimators. Composition mechanisms can
be used to combine typical machine learning workflows into a single object which
is itself an estimator, and can be used wherever usual estimators can be used.
In particular,
they integrate within model selection routines, allowing one to optimize at once
every step of a complex workflow. Composition of estimators can be done in two
ways: either sequentially through \texttt{Pipeline} objects or in a parallel
fashion through \texttt{FeatureUnion} objects.

@@

As introduced in Section~\ref{sec:estimators}, hyper-parameters set in the
constructor of an estimator
determine the behavior of its learning algorithm
and hence the performance of the resulting model on unseen data.
The problem of \textit{model selection} is therefore to find, within
some hyper-parameter space, the best combination of hyper-parameters, with
respect to some user-chosen criterion. For example, in the case of a decision
tree, too small values of the \texttt{max\_depth} parameter will make the tree
underfit, while too large values will make it overfit.

@@

Our implementation guidelines emphasize writing efficient but readable code.
In particular, we focus on making the codebase easily maintainable and
understandable in order to favor external contributions. Whenever practicable,
algorithms implemented in \sklearn are written in Python,
using NumPy vector operations for numerical work.
This allows for the code to remain concise, readable and
efficient. For critical algorithms that cannot be easily and efficiently
expressed as NumPy operations, we rely on Cython \citep{behnel2011cython}
to achieve decent performance and scalability. Cython is a
compiled programming language that extends Python with static typing. It
produces efficient C extension modules that are directly importable from the
Python run-time system. Examples of algorithms written in Cython include
stochastic gradient descent for linear models, some graph-based clustering
algorithms and decision trees.

>>>>

\begin{description}
  \item[Consistency.]
       All objects (basic or composite) share a consistent interface composed of
       a limited set of methods. This interface is documented in a consistent
       manner for all objects.
  \item[Inspection.]
       Constructor parameters and parameter values determined by learning
       algorithms are stored and exposed as public attributes.
  \item[Non-proliferation of classes.]
       Learning algorithms are the only objects to be of custom classes.
       Datasets and hyper-parameter names and values are represented as standard
       Python, NumPy or SciPy classes rather than new ones whenever feasible.
       This keeps \sklearn easy to use and easy to combine with other libraries.
  \item[Composition.]
       Many machine learning tasks are expressible
       as sequences or combinations of transformations to data.
       Some learning algorithms are also naturally viewed
       as meta-algorithms parametrized on other algorithms.
       Whenever possible, such algorithms are implemented and composed from
       existing building blocks.
  \item[Sensible defaults.]
       Whenever an operation requires a user-defined parameter,
       an appropriate default value is defined by the library.
       The default value should cause the operation to be performed
       in a sensible way (giving a baseline solution for the task at hand).
\end{description}

@@

In \sklearn, we chose a representation of data that is as close as
possible to the matrix representation: datasets are encoded as NumPy
multidimensional arrays for dense data and as SciPy sparse matrices for sparse
data. While these may seem rather unsophisticated data representations when
compared to more object-oriented constructs, such as the ones used by
Weka \citep{hall2009weka}, they bring the prime advantage of allowing us to rely
on efficient NumPy and SciPy vectorized operations while keeping
the code short and readable.  This design choice has also been motivated by
the fact that, given their pervasiveness in many other scientific Python
packages, many scientific users of Python are already familiar with NumPy dense
arrays and SciPy sparse matrices.
From a practical point of view, these formats also provide a collection of
data loading and conversion tools which make them very easy to use in many
contexts. Moreover, for tasks where the inputs are text files or semi-structured
objects, we provide \textit{vectorizer} objects that efficiently convert such
data to the NumPy or SciPy formats.

@@

The \textit{estimator} interface is at the core of the
library. It defines instantiation mechanisms of objects and exposes a
\texttt{fit} method for learning a model from training data.  All supervised and
unsupervised learning algorithms (e.g., for classification, regression or
clustering) are offered as objects implementing this interface. Machine
learning tasks like feature extraction, feature selection or dimensionality
reduction are also provided as estimators.

@@

clf = LogisticRegression(penalty="l1")
clf.fit(X_train, y_train)
\end{pythoncode}
In this snippet, a \texttt{LogisticRegression} estimator is first initialized by
setting the \texttt{penalty} hyper-parameter to \texttt{"l1"}, hence enabling
$\ell_1$ regularization. Other hyper-parameters (such as \texttt{C}, which
defines the strength of the regularization) are not explicitly defined and are
thus set to the default values. Upon calling \texttt{fit}, the model is then
learned from the training arrays \texttt{X\_train} and \texttt{y\_train},
and stored within the object for later use.
%% (In this example, \texttt{fit} is called solely for its side effects.)
Since all estimators share the same interface, using a different learning algorithm is
as simple as replacing the constructor (the class name).
For example, to build a random forest on
the same training data, one would simply have to replace
\texttt{LogisticRegression(penalty="l1")} in the snippet above by
\texttt{RandomForestClassifier()}.

In \sklearn, classical learning algorithms are not the only objects to be
implemented as estimators. For example, preprocessing routines (e.g., scaling of
features) or feature extraction techniques (e.g., vectorization of text
documents) also implement the \textit{estimator} interface. Even processing steps that are
entirely stateless and therefore do not require the \texttt{fit} method to
perform any useful work implement the estimator interface. As we will illustrate
in the next sections, this design pattern is indeed of prime importance for
consistency, composition and model selection reasons.

@@

Building on the core interface introduced in the previous section, we now
present advanced API mechanisms for building meta-estimators,
composing complex estimators and selecting models. We also discuss design
choices allowing for an easy use and extension of \sklearn.

@@

A distinguishing feature of the \sklearn API is its ability to
compose new estimators from several base estimators. Composition mechanisms can
be used to combine typical machine learning workflows into a single object which
is itself an estimator, and can be employed wherever usual estimators can be used.
In particular,
they integrate within model selection routines, allowing one to optimize at once
every step of a complex workflow. Composition of estimators can be done in two
ways: either sequentially through \texttt{Pipeline} objects or in a parallel
fashion through \texttt{FeatureUnion} objects.

@@

As introduced in Section~\ref{sec:estimators}, hyper-parameters set in the
constructor of an estimator
determine the behavior of the learning algorithm
and hence the performance of the resulting model on unseen data.
The problem of \textit{model selection} is therefore to find, within
some hyper-parameter space, the best combination of hyper-parameters, with
respect to some user-specified criterion. For example, in the case of a decision
tree, too small values of the \texttt{max\_depth} parameter will make the tree
underfit, while too large values will make it overfit.

@@

Our implementation guidelines emphasize writing efficient but readable code.
In particular, we focus on making the codebase easily maintainable and
understandable in order to favor external contributions. Whenever practicable,
algorithms implemented in \sklearn are written in Python,
using NumPy vector operations for numerical work.
This allows for the code to remain concise, readable and
efficient. For critical algorithms that cannot be easily and efficiently
expressed as NumPy operations, we rely on Cython \citep{behnel2011cython}
to achieve competative performance and scalability. Cython is a
compiled programming language that extends Python with static typing. It
produces efficient C extension modules that are directly importable from the
Python run-time system. Examples of algorithms written in Cython include
stochastic gradient descent for linear models, some graph-based clustering
algorithms and decision trees.
